Dependently Typed Languages in Statix

### 2022-10-26 paper-evcs/paper.tex
Add a few more sentences
Improve syntax of inference rules

% Commands used
\newcommand{\bhr}{\underset{\beta h}{\Rightarrow}}

@@

Note that the evaluator uses names rather than De Bruijn indices. In conventional dependently typed languages, evaluation is often done using De Bruijn indices. However, since the aim of this paper is to use the features of Statix, and scope graphs work based on names, we chose to use names.

We need to define multiple relations that will be used later for type-checking. First, the primary relation is \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression. The scope acts as the environment from \cite{krivine} paper, using \verb|NSubst| to store substitutions. All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}. We use the syntax $s1 \vdash e1, t \bhr s2 \vdash e2$ to express \verb|betaReduceHead((s1, e1)) == (s2, e2)|. Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language. One relation that is used for this is the \verb|rebuild| relation, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s.

@@

\begin{figure}[h]
	\begin{mathpar}	
		\inferrule{ 
		} {s \vdash Type(), [] \bhr s \vdash Type()}

		\inferrule{ 
			sPutSubst(s, n, (s, v)) \vdash b, t \bhr s' \vdash e'
		}{s \vdash Let(n, v, b), t \bhr s' \vdash e' }

		\inferrule{ 
			sGetName(s, n) = NSubst(se, e) 
			\\ se \vdash e, t \bhr se' \vdash  e' 
		}{ s \vdash Var(n), t \bhr se' \vdash e' }

		\inferrule{ 
			sGetName(s, n) = NType(t) 
		}{ s \vdash Var(n), t \bhr rebuild(s, Var(n), t)  }

		\inferrule{ 
		} { s \vdash FnType(n, a, b), [] \bhr s \vdash FnType(n, a, b) }

		\inferrule{ 
		} { s \vdash FnConstruct(n, a, b), [] \bhr s \vdash FnConstruct(n, a, b) }

		\inferrule{ 
			sPutSubst(s, n, a) \vdash b, ts \bhr s' \vdash e'
		} { s \vdash FnConstruct(n, a, b), t :: ts \bhr s' \vdash e' }

		\inferrule{ 
			s \vdash f, a :: ts \bhr s' \vdash e'
		} { s \vdash FnDestruct(f, a), ts \bhr s' \vdash e' }

	\end{mathpar}
	\caption{Rules for beta head reducing the Calculus of Constructions}
	\label{fig:beta-head-reduce-rules}
\end{figure}

@@

\begin{lstlisting}
typeOfExpr : scope * Expr -> Expr
\end{lstlisting}
We can then start defining type-checking rules for the language. We introduce a number of judgements for typing and equality together with their counterparts in Statix.
\begin{enumerate}
	\item $s \vdash e : t$ is the same as \verb|typeOfExpr(s, e) == t|
	\item $s1 \vdash e1 \underset{\beta}{=} s2 \vdash e2$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
	\item $s1 \vdash e1 \bhr s2 \vdash e2$ is the same as \verb|betaReduceHead((s1, e1)) == (s2, e2)|
	\item $s1 \vdash e1 \underset{\beta}{\Rightarrow} e2$ is the same as \verb|betaReduce((s1, e1)) == e2|
	\item $sEmpty \vdash e$ is the same as $e$ (empty scopes can be left out)
\end{enumerate}

@@

\begin{figure}[h]
	\begin{mathpar}	
		\inferrule{ 
		} {s \vdash Type() : Type()}

		\inferrule{ 
			s \vdash v : vt
			\\sPutSubst(s, n, (s, v)) \vdash b : t 
		}{s \vdash Let(n, v, b) : t }

@@

		\inferrule{ 
			sGetName(s, n) = NType(t) 
		}{ s \vdash Var(n) : t }

		\inferrule{ 
			sGetName(s, n) = NSubst(se, e) 
			\\ se \vdash e : t 
		}{ s \vdash Var(n) : t }

		\inferrule{ 
			s \vdash a : at 
			\\ at \underset{\beta}{=} Type() 
			\\ s \vdash a \underset{\beta}{\Rightarrow} a' 
			\\\\ sPutType(s, n, a') \vdash b : bt 
			\\ bt \underset{\beta}{=} Type() 
		} { s \vdash FnType(n, a, b) : Type() }

		\inferrule{ 
			s \vdash a : at 
			\\ at \underset{\beta}{=} Type() 
			\\ s \vdash a \underset{\beta}{\Rightarrow} a' 
			\\\\ sPutType(s, n, a') \vdash b : bt
		} { s \vdash FnConstruct(n, a, b) : FnType(n, a', bt) }

		\inferrule{ 
			s \vdash f : ft 
			\\ s \vdash ft \bhr s' \vdash FnType(da, dt, db) 
			\\\\ s \vdash a : at 
			\\ at \underset{\beta}{=} sf \vdash dt
			\\ s' \vdash db \underset{\beta}{\Rightarrow} db'
		} { s \vdash FnDestruct(f, a) : db' }

>>>>

% Commands used
\newcommand{\scope}[2]{\langle#1 \: | \: #2\rangle}
\newcommand{\bhr}[3]{ #1 \: #2 \underset{\beta h}{\Rightarrow} #3 }
\newcommand{\toe}[3]{ \scope{#1}{#2} : #3 }
\newcommand{\bred}[3]{ \scope{#1}{#2} \underset{\beta}{\Rightarrow} #3 }
\newcommand{\beq}[2]{ #1 \underset{\beta}{=} #2 }

@@

Note that the evaluator uses names rather than De Bruijn indices. In conventional dependently typed languages, evaluation is often done using De Bruijn indices. However, since the aim of this paper is to use the features of Statix, and scope graphs work based on names, we chose to use names. It would also stop us from using editor services that rely on \verb|.ref| annotations, such as renaming.

We need to define multiple relations that will be used later for type-checking. First, the primary relation is \verb|betaReduceHead|, that takes a scoped expression and a stack of applications, and returns a head-normal expression. The scope acts as the environment from \cite{krivine} paper, using \verb|NSubst| to store substitutions. All rules for \verb|betaReduceHead| are given in figure \ref{fig:beta-head-reduce-rules}. We use the syntax $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ to express \verb|betaReduceHead((s1, e1), t) == (s2, e2)|. Figure \ref{fig:beta-head-reduce-rules} contains the rules necessary for beta head reduction of the language. One relation that is used for this is the \verb|rebuild| relation, which takes a scoped expression and a list of arguments and converts it to an expression by adding \verb|FnDestruct|s.

@@

\begin{figure}[h]
	\begin{mathpar}	
		% Type rule
		\inferrule{ 
		} { 
			\bhr
			{ \scope{s}{Type} }
			{ [] }
			{ \scope{s}{Type}} 
		}

		% Let rule
		\inferrule{ 
			\bhr 
			{ \scope{ sPutSubst(s, n, (s, v))}{b} } 
			{ t }
			{ \scope{ s' } { e' } }
		}{
			\bhr
			{ \scope{ s }{ Let(n, v, b) } }
			{ t }
			{ \scope { s' } { e' } }
		}

		% Var rule - NSubst
		\inferrule{ 
			sGetName(s, n) = NSubst(se, e) 
			\\ \bhr
			{\scope{se}{e}}
			{t}
			{\scope{se'}{e'}} 
		}{
			\bhr
			{\scope{s}{Var(n)}}
			{t}
			{\scope{se'}{e'}} 
		}

		% Var rule - NType
		\inferrule{ 
			sGetName(s, n) = NType(t) 
		}{ 
			\bhr
			{\scope{s}{Var(n)}}
			{t}
			{rebuild(s, Var(n), t)} 
		}

		% FnType rule
		\inferrule{ 
		} { 
			\bhr
			{\scope{s}{FnType(n, a, b)}}
			{[]}
			{\scope{s}{FnType(n, a, b)}} 
		}

		% FnConstruct rule - No args
		\inferrule{ 
		} { 
			\bhr
			{\scope{s}{FnConstruct(n, a, b)}}
			{[]}
			{\scope{s}{FnConstruct(n, a, b)}} 
		}

		% FnConstruct rule - Args
		\inferrule{ 
			\bhr
			{\scope{sPutSubst(s, n, a)}{b}}
			{ts}
			{\scope{s'}{e'}} 
		} { 
			\bhr
			{\scope{s}{FnConstruct(n, a, b)}}
			{(t::ts)}
			{\scope{s'}{e'}} 
		}

		% FnDestruct rule
		\inferrule{ 
			\bhr
			{\scope{s}{f}}
			{(a::ts)}
			{\scope{s'}{e'}} 
		} { 
			\bhr
			{\scope{s}{FnDestruct(f, a)}}
			{ts}
			{\scope{s'}{e'}} 
		}
	\end{mathpar}
	\caption{Rules for beta head reducing the Calculus of Constructions}
	\label{fig:beta-head-reduce-rules}
\end{figure}

@@

\begin{lstlisting}
typeOfExpr : scope * Expr -> Expr
\end{lstlisting}
We can then start defining type-checking rules for the language. We introduce a number of judgements for typing and equality together with their counterparts in Statix.
\begin{enumerate}
	\item $\toe{s}{e}{t}$ is the same as \verb|typeOfExpr(s, e) == t|
	\item $\beq{\scope{s1}{e1}}{\scope{s2}{e2}}$ is the same as \verb|expectBetaEq((s1, e1), (s2, e2))|
	\item $\bhr{\scope{s1}{e1}}{t}{\scope{s2}{e2}}$ is the same as \verb|betaReduceHead((s1, e1)) == (s2, e2)| \\ (The same as in section \ref{sec:coc-dynsyms})
	\item $\bred{s1}{e1}{e2}$ is the same as \verb|betaReduce((s1, e1)) == e2|
	\item $\scope{sEmpty}{e}$ is the same as $e$ (empty scopes can be left out)
\end{enumerate}

@@

\begin{figure}[h]
	\begin{mathpar}	
		% Type rule
		\inferrule{ 
		} {
			\toe{s}{Type()}{Type()}
		}

		% Let rule
		\inferrule{ 
			\toe{s}{v}{vt}
			\\ \toe{sPutSubst(s, n, (s, v))}{b}{t}
		}{
			\toe{s}{Let(n, v, b)}{t}
		}

@@

		% Var rule - NType
		\inferrule{ 
			sGetName(s, n) = NType(t) 
		}{ 
			\toe{s}{Var(n)}{t}
		}

		% Var rule - NSubst
		\inferrule{ 
			sGetName(s, n) = NSubst(se, e) 
			\\ \toe{se}{e}{t}
		}{
		 	\toe{s}{Var(n)}{t}
		}

		% FnType rule
		\inferrule{ 
			\toe{s}{a}{at}
			\\ \beq{at}{Type()} 
			\\ \bred{s}{a}{a'}
			\\\\  \toe{sPutType(s, n, a')}{b}{bt}
			\\ \beq{bt}{Type()}
		} { 
			\toe{s}{FnType(n, a, b)}{Type()}
		}

		% Force these to be on the same line, using negative spaces
		\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!

		% FnConstruct rule
		\inferrule{ 
			\toe{s}{a}{at}
			\\ \beq{at}{Type()}
			\\ \bred{s}{a}{a'} 
			\\\\ \toe{sPutType(s, n, a')}{b}{bt}
		} { 
			\toe{s}{FnConstruct(n, a, b)}{FnType(n, a', bt)}
		}

		% FnDestruct rule
		\inferrule{ 
			\toe{s}{f}{ft}
			\\ \bhr
				{\scope{s}{ft}}
				{[]}
				{\scope{s'}{FnType(da, dt, db)}} 
			\\\\ \toe{s}{a}{at} 
			\\ \beq{at}{\scope{sf}{dt}}
			\\ \bred{s'}{db}{db'}
		} { 
			\toe{s}{FnDestruct(f, a)}{db'}
	 	}
